Is high social class always beneficial for survival? A study of northern Sweden 1801--2013 \footnote{Short; "Is high social class beneficial for survival?"} \footnote{Funding:This work was founded by the Swedish Research Council under grants No 349-2008-6592 No 421-2012-892}

### 2018-10-01 popstud_open.Rmd
Some final corrections

---
title: Is high social class always beneficial for survival? A study of northern Sweden 1801--2013 \footnote{Short; "Is high social class beneficial for survival?"} \footnote{Funding:}
author: Sören Edvinsson \footnote{email soren.edvinsson@umu.se, orcid 0000-0001-7439-002X (corresponding author)}
  and Göran Broström \footnote{email goran.brostrom@umu.se, orcid 0000-0002-2865-7894}
date: Centre for Demographic and Ageing Research, Umeå University, Umeå, Sweden
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: yes
    toc_depth: 2
citation_package: natbib
classoption: titlepage
biblio-style: apalike
documentclass: article
fontsize: 12pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{endfloat}
- \usepackage[doublespacing]{setspace}
- \usepackage{sectsty}
- \subsectionfont{\normalfont\itshape}
bibliography: mybib
abstract: "Focusing on two regions in northern Sweden 1801--2013, we challenge common
  notions of the assumed advantage in survival of belonging to a high social class.
  The issue is analysed according to gender and age group (adults and elderly) and
  in relation to the development of economic inequality. The results show that high
  social class is not always favourable for survival. Men in the elite category, particularly
  in working age, had higher mortality compared to others during a large part of the
  studied period; a male mortality class reversal appears at a surprisingly late date,
  while the social gradient among women conforms to the expected pattern. We suggest
  that health-related behaviour is decisive not only in later but earlier phases of
  the mortality transition as well. The results implicate that the association between
  social class and health is more complex than is assumed in many of the dominant
    theories in demography and epidemiology.\\par\n\\textbf{Keywords:} Class reversal
  in mortality; Cox regression; Gender differences; Inverse probability weighting;
  Mortality transition; Northern Sweden; Social inequality; 19th to 21 century."

@@

Figure \@ref(fig:plottaalla) shows the distribution
(per cent) of exposure time according to class for men and women,
respectively. The farmer category was dominant until the middle of the 20th centuries but have become marginal in the latest periods, particularly among women. The category of workers is fairly stable over time. The highest class has increased substantially, albeit from a very low level
in the 19th century. The increase in academics in connection to the
establishment of Umeå University explains much of this increase in later time. Furthermore,
we observe a rather high percentage of missing social class in the early 19th century.
For men the proportion is somewhat higher in the age group 65 and above, those who were retired,
but still low enough to be used in our analyses (results not shown). If we disregard the very first period,
however, there were very few with occupation missing until 1950.

@@

The proportion with missing values among women is a larger problem, however;
see Figure \@ref(fig:plottaalla). Corresponding to what we found for men, the
proportion without social position is very low until the middle of the 20th
century, reflecting the way women's social position is defined. However, the frequency of missing social class is quite high during particularly the first two periods of data from the Linnaeus database---from 1976. Since female social class for this period is defined
according to own occupation, the consequence of many women being outside
the workforce is that the proportion of missing increases. For the later periods,
the results are fairly complete.

@@

Figure 3 is shown as absolute numbers, exposures for women in Table \@ref(tab:womenexp), and exposures for men in Table \@ref(tab:menexp).

@@

Then number of deaths for women are shown in Table \@ref(tab:womendeaths), and for men in Table \@ref(tab:mendeaths).

>>>>

---
title: Is high social class always beneficial for survival? A study of northern Sweden 1801--2013 \footnote{Short; "Is high social class beneficial for survival?"} \footnote{Funding:This work was founded by the Swedish Research Council under grants No 349-2008-6592 No 421-2012-892}
author: Sören Edvinsson \footnote{email soren.edvinsson@umu.se, orcid 0000-0001-7439-002X (corresponding author)}
  and Göran Broström \footnote{email goran.brostrom@umu.se, orcid 0000-0002-2865-7894}
date: Centre for Demographic and Ageing Research, Umeå University, Umeå, Sweden
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: yes
    toc_depth: 2
citation_package: natbib
classoption: titlepage
biblio-style: apalike
documentclass: article
fontsize: 12pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{endfloat}
- \usepackage[doublespacing]{setspace}
- \usepackage{sectsty}
- \subsectionfont{\normalfont\itshape}
bibliography: mybib
abstract: "Focusing on two regions in northern Sweden 1801--2013, we challenge common
  notions of the assumed advantage in survival of belonging to a high social class.
  The issue is analysed according to gender and age group (adults and elderly) and
  in relation to the development of economic inequality. The results show that high
  social class is not always favourable for survival. Men in the elite category, particularly
  in working age, had higher mortality compared to others during a large part of the
  studied period; a male mortality class reversal appears at a surprisingly late date,
  while the social gradient among women conforms to the expected pattern. We suggest
  that health-related behaviour is decisive not only in later but earlier phases of
  the mortality transition as well. The results implicate that the association between
  social class and health is more complex than is assumed in many of the dominant
    theories in demography and epidemiology.\\par\n\\textbf{Keywords:} Class reversal
  in mortality; Cox regression; Gender differences; Inverse probability weighting;
  Mortality transition; Northern Sweden; Social inequality; 19th to 21 century."

@@

Figure \@ref(fig:plottaalla) shows the distribution
(per cent) of exposure time according to class for men and women,
respectively (see also Appendix). The farmer category was dominant until the middle of the 20th centuries but have become marginal in the latest periods, particularly among women. The category of workers is fairly stable over time. The highest class has increased substantially, albeit from a very low level
in the 19th century. The increase in academics in connection to the
establishment of Umeå University explains much of this increase in later time. Furthermore,
we observe a rather high percentage of missing social class in the early 19th century.
For men the proportion is somewhat higher in the age group 65 and above, those who were retired,
but still low enough to be used in our analyses (results not shown). If we disregard the very first period,
however, there were very few with occupation missing until 1950.

@@

The proportion with missing values among women is a larger problem, however. Corresponding to what we found for men, the
proportion without social position is very low until the middle of the 20th
century, reflecting the way women's social position is defined. However, the frequency of missing social class is quite high during particularly the first two periods of data from the Linnaeus database---from 1976. Since female social class for this period is defined
according to own occupation, the consequence of many women being outside
the workforce is that the proportion of missing increases. For the later periods,
the results are fairly complete.

@@

The exposure in years for men and women respectively are shown in Table \@ref(tab:womenexp), and Table \@ref(tab:menexp).

@@

The number of deaths for women are shown in Table \@ref(tab:womendeaths), and for men in Table \@ref(tab:mendeaths).
